---
title: "WHO COVID-19 Data Analysis: Pre-existing Conditions & Policy Changes"
date: "`r Sys.Date()`"
author: "Rahul Uppuluri, Jennifer Li, Richard Ge"
output: html_document
---

<!--
To update Github with your own code: 
1. PULL
From the Git tab, “Pull” the repository. This makes sure your local repository is synced with the remote repository. This is very important if other people are making changes to the repository or if you are working from multiple computers. 
2. STAGE
Select (aka. stage) the files you want to commit. In RStudio, this involves checking the “Staged” boxes in the Git tab. 
3. COMMIT
Try to add a note so everyone will know what you're changing. 
4. PUSH
Saves these local changes to the online repository.
-->
<!--
## Overview

This document contains instructions on Project 2 for STA 141A in Winter 2021. This document is made with `R markdown`. The `rmd` file to generate this document is available on the course website.

## Background



## WHO COVID-19 data

We will do an EDA for the WHO COVID-19 data. You can take a look at the [weekly WHO COVID-19 update](https://www.who.int/emergencies/diseases/novel-coronavirus-2019/situation-reports) for reference.


```{r echo=TRUE, eval=TRUE}
sessionInfo()
# library(tidyverse)
# covid <- read_csv("https://covid19.who.int/WHO-COVID-19-global-data.csv")
```

This data set is maintained by WHO and updated constantly. The first task for you is to understand this data set, e.g., the meaning of the variables and their values. To this end, you can make use of your preferred search engine, or read the documentation for this data set. 

In this project, you are a team of conscientious statisticians, who wish to help the general public understand the ongoing pandemic. 
-->

<!--
## Suggested outline 

The following list provides one potential structure of the data analysis report. As this is the final project, the following suggestions are intended to provide one viable route for your project while leaving you as much freedom as possible. 
-->
<!--
***

0. Before writing your analysis report, you may want to explore this data set and read about the coronavirus to generate the hypothesis or question to be answered in this report, i.e., the question(s) of interest. You can be creative on this question so long as it meets three conditions.

    a. This question might be meaningful in real-world. 
    b. A method discussed in this course is employed in solving this question. 
    c. This question does not ask for causality.  
-->


***

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#### SETUP
# install.packages("usethis")
library(usethis, warn.conflicts=F)
    use_git_config(user.name = "Richard-Ge", user.email = "richardge951@gmail.com")
library(tidyverse, warn.conflicts=F)
library(pastecs)
library(car, warn.conflicts=F)
    covid <- read_csv("https://covid19.who.int/WHO-COVID-19-global-data.csv")
library(lubridate, warn.conflicts=F)
# install.packages("rvest")
library(rvest, warn.conflicts=F)
library(data.table, warn.conflicts=F) 
```

## 1. Dataset, Questions of Interest
<!--Introduce the dataset and the questions of interest. -->
The global COVID-19 pandemic has caused a large number of deaths since it was first discovered in 2019. Here, we will be exploring if existing (pre-2020) markers for each country, such as air pollution levels, healthcare quality, and average population age, are associated with a higher death toll, and if policy changes in government are associated with fewer deaths. 

We will primarily be using the WHO's global COVID-19 dataset in this report, which includes the date a country (labeled by their country code and also by their region in the world) reported a new case or death, as well as the updated cumulative number of cases and deaths. We will only be exploring specific Eastern European countries compared to specific Western European countries in this report because of their similar yet vastly different environments--similar weather, social healthcare provisions, and government policies vs. differing air pollution levels and population ages. 

Our second question of interest is examining how policy updates (mask mandates and stay-at-home restrictions) in those countries affect the number of new cases per capita from March to October. Since the levels of each of these policies may have different meanings in different countries, we will only be exploring one specific European country (France) that displays a strong balance between all of the categories for this analysis (i.e. recommended, required with exceptions, etc.). Additionally, we are focusing on the specific timeframe of March to October to see how policies normalized overtime, as March marked the beginning of the lockdown period in which many policies were enacted in countries, while October is the six-month milestone before the major Holiday season.

## 2. Background
<!--Review the background of the coronavirus. -->
The coronavirus disease 2019, also known as COVID-19, is a virus known to cause mild to severe respiratory illness, and originated from an outbreak in Wuhan, China. This highly contagious virus spread throughout the world very quickly, and the CDC declared it a pandemic due to its deadly nature in March. In order to take preventative measures from catching it, many schools and businesses were shut down globally as the world moved to online learning or remote work from home (unless deemed essential by their government), and social guidelines or rules were enacted, which include social distancing, wearing masks, self-isolation, and other hygienic practices. While people waited for more research on how the virus spread and for a vaccine to be developed and distributed, many nations across the world faced high death tolls and rising case numbers.

## 3. Summary Statistics 
<!--
Explore this dataset and generate summary statistics that you find informative, and explain your findings. <span style="color:red"> The summary statistics should include at least time, number of cases, number of death, case-mortality rate.</span>
-->

### Linear Model

It has been 425 days since the WHO began collecting data, and 404 days since the first COVID-19 cases appeared in France. There have been a total of 33,240,905 cases and 765,170 deaths in the European countries selected for this analysis, with a mortality rate of 2.30%. The mean of percent changes in case numbers (0.575%) is over 200 times larger than the median (0.00252%), which indicates that the increases in case numbers are strongly skewed right; on some days, the number of cases (adjusted for population) in some countries increases very quickly compared to the other days. 

When accounting for whether our selected countries are in the EU, the EU countries have 22,478,344 cases and 550,689 deaths, while the non-EU countries have 10,675,458 cases and 212,843 deaths. This difference in case numbers is most likely due to the greater number of, and higher population in, EU countries as opposed to the fewer, less-populated European countries in this analysis. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# head(covid) # View(covid)
covid = mutate(covid, WHO_region = fct_recode(WHO_region,"Eastern Mediterranean"="EMRO","Europe" = "EURO","Africa" = "AFRO","Western Pacific" = "WPRO","Americas"="AMRO","South-East Asia" = "SEARO"))

## ADDING DATASETS
pop.data = read.csv("~/Documents/STA141A-pop-data.csv", row.names=1)  # View(pop.data)
obesity = read_tsv("~/Documents/obesity.tsv")
    obesity = obesity %>% transmute(Country=obesity$Country,ObesityPrc=obesity$Obesity_rate) %>% data.frame()
elderly = read_tsv("~/Documents/elderly2017WB.tsv")
    elderly = elderly %>% transmute(Country=elderly$Country,ElderlyPrc=elderly$`c-65+`) %>% data.frame()
healthcare = read_tsv("~/Documents/healthcare.tsv")
    healthcare = healthcare %>% transmute(Country=healthcare$Country,HCSpentPerCap=healthcare$pc2018) %>% data.frame()
aqis = read_tsv("~/Documents/country-aqi.tsv")
    # no need to transmute
pop_dens = read_tsv("~/Documents/pop_dens.tsv")
    pop_dens = pop_dens %>% transmute(Country=pop_dens$Country,km.density=pop_dens$km_PopDens) %>% data.frame()
incomes = read_tsv("~/Documents/incomes.tsv")
    incomes = incomes %>% transmute(Country=incomes$Country,med.wealth=incomes$MedWealth) %>% data.frame()
HDIs = read_tsv("~/Documents/countryHDIs.tsv")
    HDIs = HDIs %>% transmute(Country=HDIs$Country,HDIadj=HDIs$HDI.ineq) %>% data.frame()

## RENAMING SOME ENTRIES (do this directly in the data files!)
pop.data[85, 1] = "Dominican Republic"
pop.data[188, 1] = "American Samoa"
pop.data[16, 1] = "Democratic Republic of the Congo"
pop.data[19, 1] = "Iran (Islamic Republic of)"
pop.data[21, 1] = "The United Kingdom"
pop.data[50, 1] = "Venezuela (Bolivarian Republic of)"
pop.data[81, 1] = "Bolivia (Plurinational State of)"
pop.data[86, 1] = "Czechia"
pop.data = rbind(pop.data, Country = c("Andorra", as.numeric(77543)))
pop.data = rbind(pop.data, Country = c("Latvia", as.numeric(1907675)))
# pop.data = rbind(pop.data, Country = c("Liechtenstein", as.numeric(38250)))
pop.data = rbind(pop.data, Country = c("Moldova", as.numeric(2640438)))
# pop.data = rbind(pop.data, Country = c("Monaco", as.numeric(38300)))
pop.data = rbind(pop.data, Country = c("Moldova", as.numeric(2640438)))
pop.data = rbind(pop.data, Country = c("Russia", as.numeric(146238185)))
pop.data = mutate(pop.data, Population = as.numeric(Population))  # ^^^ gets inputted as char for some reason
obesity[46, 1] = "Czechia"
obesity[36, 1] = "The United Kingdom"
obesity = mutate(obesity, ObesityPrc = as.numeric(ObesityPrc))
elderly = rbind(elderly, Country = c("Andorra", as.numeric(17.4)))
elderly[46, 1] = "Czechia"
elderly[182, 1] = "The United Kingdom"
elderly = mutate(elderly, ElderlyPrc = as.numeric(ElderlyPrc))
# for the healthcare data, i realized that i could just edit the tsv itself...no need for all those reassignments...

## JOINING NEW DATA TO WHO DATASET
newcovid = left_join(covid, pop.data, by = c("Country"))
newcovid = left_join(newcovid, obesity, by = c("Country"))
newcovid = left_join(newcovid, elderly, by = c("Country"))
newcovid = left_join(newcovid, healthcare, by = c("Country"))
newcovid = left_join(newcovid, aqis, by = c("Country"))
newcovid = left_join(newcovid, pop_dens, by = c("Country"))
newcovid = left_join(newcovid, incomes, by = c("Country"))
newcovid = left_join(newcovid, HDIs, by = c("Country"))

    # put new populations here, by name
# View(newcovid); View(pop.data)   # got confused: saw the same pop. many times
newcovid = filter(newcovid, !is.na(newcovid$Population))  # REMOVE NA pop. countries
worldpop = sum(unique(newcovid$Population), na.rm = T)     # having NA values makes output NA too
## ADD EXTRA STATISTICS (these vary too much; ended up only using mortality.now)
newcovid = mutate(newcovid, 
                  prcCNew=New_cases*100/Population,       # new cases/pop -> %
                  prcCTotal=Cumulative_cases*100/Population, # total cases/pop -> %
                  prcDNew=New_deaths*100/Population,   # new deaths/pop -> %
                  prcDTotal=Cumulative_deaths*100/Population, # total deaths/pop -> %
                  mortality.now=Cumulative_deaths/Cumulative_cases)  
# View(newcovid)
## TODO: get summary statistics e.g. total deaths (+ per region), percent of country infected, etc.?

#covid20 = newcovid %>% filter(Date_reported < "2021-01-01")  # View(covid20)
#covid21 = newcovid %>% filter(Date_reported >= "2021-01-01")

## CREATE OUR EUROPE DATASET
time.since.start = max(covid$Date_reported) - min(covid$Date_reported)
time.since.first = max(covid$Date_reported) - ymd("2020-1-24")
recent.data = filter(covid, Date_reported == Sys.Date()-1)  # or today()
# creating dataset for "europe pre-existing markers" model:
EUcountries = c("Austria", "Belgium", "Bulgaria", "Croatia", "Cyprus", "Czechia", "Denmark", "Estonia", "Finland", "France", "Germany", "Greece", "Hungary", "Ireland", "Italy", "Latvia", "Lithuania", "Luxembourg", "Malta", "Netherlands", "Poland", "Portugal", "Romania", "Slovakia", "Slovenia", "Spain", "Sweden")
nonEUcountries = c("Albania", "Andorra", "Armenia", "Azerbaijan", "Belarus", "Bosnia and Herzegovina", "Georgia", "Iceland", "Liechtenstein", "Moldova", "Monaco", "Montenegro", "Norway", "Russia", "Serbia", "Switzerland", "Turkey", "Ukraine", "The United Kingdom")
europecovid = filter(newcovid, newcovid$Country %in% c(EUcountries, nonEUcountries))
 # sort(c(EUcountries, nonEUcountries)); sort(unique(europecovid$Country))
europe.recent = filter(europecovid, Date_reported == Sys.Date()-1)  # View(recent.europe)
    # latvia is reported twice?
# mutate(europecovid, Population = )


## SUMMARY STATS: include at least time, number of cases, number of death, case-mortality rate.
eu.total.cases = sum(europe.recent$Cumulative_cases)
eu.total.deaths = sum(europe.recent$Cumulative_deaths)
eu.total.mortality = eu.total.deaths*100/eu.total.cases
eu.mean.cases = sum(europecovid$prcCNew)/as.numeric(time.since.first)
eu.median.cases = median(europecovid$prcCNew)
eu.stats = data.frame(time.since.start, time.since.first, eu.total.cases, eu.total.deaths, eu.total.mortality, eu.mean.cases, eu.median.cases)
names(eu.stats) = c("Days since WHO data began", "Days since first reported case", "Total Cases", "Total Deaths", "Total Mortality (as %)", "Mean of % Changes in Case Numbers", "Median of % Changes in Case Numbers")
# View(europe.recent); View(pop.data); View(covid21)
### time since first case, cases, prc. of total cases, deaths, prc. of total deaths, mortality rate
# time.since.first.case = c(0, 0, 0)
cases = c(0, 0)
# prc.total.cases = c(0, 0, 0)
deaths = c(0, 0)
# prc.total.deaths = c(0, 0, 0)
# mortality.rate = c(0, 0, 0)
summarystats = data.frame(cases, deaths, row.names=c("Selected EU countries", "Selected Non-EU Countries"))
europop = sum(europe.recent$Population)
for(country1 in unique(europecovid$Country)){
  # print(country)
  if((country1 %in% EUcountries)) {
    thiscountry = filter(europe.recent, Country==country1)
    summarystats[1,1] = summarystats[1,1] + unique(thiscountry$Cumulative_cases)
    #summarystats[1,2] = summarystats[1,2] + unique(thiscountry$Cumulative_cases)*100/europop
    summarystats[1,2] = summarystats[1,2] + unique(thiscountry$Cumulative_deaths)
    #summarystats[1,4] = summarystats[1,4] + unique(thiscountry$Cumulative_deaths)*100/europop
    # summarystats[1,5] = summarystats[1,5] + unique(thiscountry$mortality.now)
    # print(cat("EU:  ", country1, unique(thiscountry$Cumulative_cases)))
  } 
  if((country1 %in% nonEUcountries)){
    thiscountry = filter(europe.recent, Country==country1)
    summarystats[2,1] = summarystats[2,1] + unique(thiscountry$Cumulative_cases)
    #summarystats[2,2] = summarystats[2,2] + unique(thiscountry$Cumulative_cases)*100/europop
    summarystats[2,2] = summarystats[2,2] + unique(thiscountry$Cumulative_deaths)
    #summarystats[2,4] = summarystats[2,4] + unique(thiscountry$Cumulative_deaths)*100/europop
    # summarystats[2,5] = summarystats[2,5] + unique(thiscountry$mortality.now)
    # print(cat("Not: ", country1, unique(thiscountry$Cumulative_cases)))
  }
}
#summarystats[3,1] = summarystats[1,1] + summarystats[2,1]
#summarystats[3,2] = (summarystats[1,1] + summarystats[2,1])*100/sum(europe.recent$Population)
#summarystats[3,2] = summarystats[1,2] + summarystats[2,2]
#summarystats[3,4] = summarystats[1,4] + summarystats[2,4]
#summarystats[3,5] = summarystats[1,5] + summarystats[2,5]
names(summarystats) = c("Total Cases", "Total Deaths")
## REPORT SUMMARY STATS:
eu.stats; summarystats
europe.recent %>% select(9:16, 21) %>% stat.desc(basic=F)
```
<!--end of Richard's summary stats-->

### ANOVA Model: Policy Summary Stats

For France, the number of new cases has a mean of 2503 and a median of 1025, indicating that the distribution is positively skewed. Out of the 215 observations of new cases for France, one of them appears to be negative at a value of -840, which seems to be an outlier but does not have any leverage on the rest of the data as indicated by the Residuals versus Leverage plot generated during sensitivity analysis later on in the report. Additionally, the variable has a lower quartile of 356, an upper quartile of 3742, and a standard deviation of 3301.58. A boxplot of each individual policy was plotted against new cases as well their counts. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
face.coverings <- read_csv("~/Documents/face-covering-policies-covid.csv")
stay.at.home <- read_csv("~/Documents/stay-at-home-covid.csv")
public.events <- read_csv("~/Documents/public-events-covid.csv")

colnames(face.coverings)[1] <- "Country"
colnames(stay.at.home)[1] <- "Country"
colnames(public.events)[1] <- "Country"

colnames(face.coverings)[3] <- "Date_reported"
colnames(stay.at.home)[3] <- "Date_reported"
colnames(public.events)[3] <- "Date_reported"

var = "France"
covid2 <- filter(covid, Country == var, Date_reported >= "2020-03-01", Date_reported <= "2020-10-01")
face.coverings <- filter(face.coverings, Country == var, Date_reported >= "2020-03-01", Date_reported <= "2020-10-01")
stay.at.home <- filter(stay.at.home, Country == var, Date_reported >= "2020-03-01", Date_reported <= "2020-10-01")
public.events <- filter(public.events, Country == var, Date_reported >= "2020-03-01", Date_reported <= "2020-10-01")

policy.data <- inner_join(covid2,face.coverings, by=c("Date_reported","Country"))
policy.data <- inner_join(policy.data, stay.at.home, by=c("Date_reported","Country"))

drops <- c("Code.x", "Code.y", "Code", "Country_code")
policy.data[ , !(names(policy.data) %in% drops)]
#head(policy.data)

"Summaries of Variables"
policy.data %>% select(5:12) %>% stat.desc(basic=F)

policy.data$facial_coverings = factor(policy.data$facial_coverings)
levels(policy.data$facial_coverings) <- list(No_policy = "0", Recommended = "1", Required_in_some = "2", Required_in_all = "3", Required_outside = "4")
#is.factor(policy.data$facial_coverings)
#head(policy.data)
#unique(policy.data$facial_coverings)

policy.data$stay_home_requirements = factor(policy.data$stay_home_requirements)
levels(policy.data$stay_home_requirements) <- list(No_measures = "0", Recommended_not_leave = "1", Required_with_exceptions = "2", Required_miimal_exceptions = "3")
#is.factor(policy.data$stay_home_requirements)
#head(policy.data)
#unique(policy.data$stay_home_requirements)

```

##### Univariate Analysis
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(pastecs)
```

```{r, echo=FALSE}
#"Summaries of Data"
#summary(policy.data) #No NaNs in dataset, so no need to drop anything
"Summary of New Cases"
summary(policy.data$New_cases)

negative.cases <- filter(policy.data, policy.data$New_cases < 0)
#nrow(negative.cases) %>% print
#nrow(policy.data) %>% print

ggplot(data=policy.data, aes(x=Date_reported, y=New_cases)) +
  geom_line(alpha=0.9) +
  ggtitle("Date vs New Cases") +
  xlab("Date") +
  ylab("New Cases")

ggplot(data=policy.data, mapping=aes(x=facial_coverings, y=New_cases)) +
  geom_boxplot() +
  ggtitle("Facial Covering Requirements vs New Cases") +
  xlab("Facial Covering Requirements") +
  ylab("New Cases")

ggplot(data=policy.data, mapping=aes(x=facial_coverings)) +
  geom_bar() +
  ggtitle("Count of Facial Covering Requirements") +
  xlab("Facial Covering Requirements") +
  ylab("Count")


ggplot(data=policy.data, mapping=aes(x=stay_home_requirements, y=New_cases)) +
  geom_boxplot() +
  ggtitle("Stay-at-home Requirements vs New Cases") +
  xlab("Stay-at-home Requirements") +
  ylab("New Cases")

ggplot(data=policy.data, mapping=aes(x=stay_home_requirements)) +
  geom_bar() +
  ggtitle("Count of Stay-at-home Requirements") +
  xlab("Stay-at-home Requirements") +
  ylab("Count")


```

##### Multivariate Analysis
```{r, echo=FALSE}
boxplot(policy.data$New_cases ~ policy.data$facial_coverings, xlab = "Facial Covering Requirements", ylab = "New Cases", main="Boxplot of New Cases by Facial Covering Requirements")

boxplot(policy.data$New_cases ~ policy.data$stay_home_requirements, xlab = "Stay-at-home Requirements", ylab = "New Cases", main="Boxplot of New Cases by Stay-at-home Requirements")

```

## 4. Models
<!--
Propose an appropriate model to answer the questions of interest. 
      a. Explain your notations. 
      b. State assumptions for your model. 
      c. Explain why your model is appropriate for this task. 
-->

### Linear Model

One model we will use is the simple linear regression model, which is defined as follows: $$Y_{i} = \beta_{0} + \beta_{1}X_{i1} + \beta_{2}X_{i2} + ... + \beta_{p}X_{ip} + \epsilon_{i}$$ 

This model is ideal for analyzing what pre-COVID metrics of a country are associated with the current mortality rate, because all of the variables are continuous. Furthermore, there is only one outcome variable, and many possible predictor variables, which are well-suited to linear dimension reduction methods. 

There are $p+1$ (when including the intercept $\beta_{0}$) parameters $\beta_{p}$, $p$ predictor variables $X_{ip}$, and outcome variables $Y_i$. The error terms are represented by $\epsilon_{i}$. The index $i$ denotes the $i$'th observation (in our context, the $i$'th country analyzed), while the index $p$ denotes the $p$'th predictor ($p$ denotes a metric for measuring countries, while the intercept of the model is at $p=0$). 

We will use mortality rate as the outcome variable $Y_i$ because it is most likely to be directly associated with how susceptible a country's population is to the pandemic, as opposed to infection rates/day-to-day statistics (which have greater variation both between and within countries) and simple cumulative statistics (which seem to only increase with time, making the analysis biased towards fewer cases/deaths).  <maybe connect sta131b concepts to this reasoning? Analysis will be biased anyways as more countries get things under control, and the mortality rate shifts>

The proposed predictors for this model are as follows: percent of the population that is obese, percent of the population that is elderly, total health expenditure per capita (in USD), air quality index (AQI) aggregated per country, population density (per km^2), median wealth in USD (per adult), and inequality-adjusted HDI (the unadjusted HDI is the theoretical development of a country, while the adjusted version accounts for a decrease due to inequality). 

The assumptions are as follows: there is a linear relationship between the predictor and outcome variables, the variables are independent from each other, there is no multicollinearity between the variables, and the residuals are linear and normally distributed. 

```{r, echo=FALSE}

groups = as.data.table(europecovid)

## THE COMMENTED CODE IS MOSTLY GARBAGE, MAY BE DELETED LATER!
    # get highest days of new/total cases and new/total deaths
#europe.highest.CNew = groups[groups[, .I[prcCNew==max(prcCNew)], by=Country]$V1]
#dim(europe.highest.CNew);length(unique(europe.highest.CNew$prcCNew)) # no repeats
#europe.highest.CNew = europe.highest.CNew %>% mutate(  #orig. HC, med.wealth only
#  HCSpentPerCap = scale(HCSpentPerCap), 
#  med.wealth=scale(med.wealth), 
#  ElderlyPrc=scale(ElderlyPrc),
#  ObesityPrc=scale(ObesityPrc),
#  AQI=scale(AQI),
#  km.density=scale(km.density),
#  prcCNew=scale(prcCNew),
#)

#europe.highest.CTotal = groups[groups[, .I[prcCTotal==max(prcCTotal)], by=Country]$V1]
#dim(europe.highest.CTotal);length(unique(europe.highest.CTotal$prcCTotal)) # a few repeats!
#europe.highest.CTotal=europe.highest.CTotal[!duplicated(europe.highest.CTotal$prcCTotal), ]  # from stackoverflow
#europe.highest.CTotal = europe.highest.CTotal %>% mutate(  #orig. HC, med.wealth only
#  HCSpentPerCap = scale(HCSpentPerCap), 
#  med.wealth=scale(med.wealth), 
#  ElderlyPrc=scale(ElderlyPrc),
#  ObesityPrc=scale(ObesityPrc),
#  AQI=scale(AQI),
#  km.density=scale(km.density),
#  prcCTotal=scale(prcCTotal),
#)

#europe.highest.DNew = groups[groups[, .I[prcDNew==max(prcDNew)], by=Country]$V1]
#dim(europe.highest.DNew);length(unique(europe.highest.DNew$prcDNew)) # a few repeats!
#europe.highest.DNew=europe.highest.DNew[!duplicated(europe.highest.DNew$prcDNew), ]
#europe.highest.DNew = europe.highest.DNew %>% mutate( 
#  HCSpentPerCap = scale(HCSpentPerCap), 
#  med.wealth=scale(med.wealth), 
#  ElderlyPrc=scale(ElderlyPrc),
#  ObesityPrc=scale(ObesityPrc),
#  AQI=scale(AQI),
#  km.density=scale(km.density),
#  prcDNew=scale(prcDNew),
#)

#europe.highest.DTotal = groups[groups[, .I[prcDTotal==unique(max(prcDTotal))], by=Country]$V1]
#dim(europe.highest.DTotal);length(unique(europe.highest.DTotal$prcDTotal)) # many repeats!
#europe.highest.DTotal=europe.highest.DTotal[!duplicated(europe.highest.DTotal$prcDTotal), ]
#europe.highest.DTotal = europe.highest.DTotal %>% mutate(  #orig. HC, med.wealth only
#  HCSpentPerCap = scale(HCSpentPerCap), 
#  med.wealth=scale(med.wealth), 
#  ElderlyPrc=scale(ElderlyPrc),
#  ObesityPrc=scale(ObesityPrc),
#  AQI=scale(AQI),
#  km.density=scale(km.density),
#  prcDTotal=scale(prcDTotal),
#)

# plot(europe.highest.DTotal$km.density)  ## big outlier: malta


# View(filter(europecovid, Country=="Finland"))     View(europe.highest)

# OLD PLANS: 
## simple linreg (+ x-var squared?) for the whole time, or for each year/month/week
  # can we do many models instead of just 1?
### we might be able to GROUP these months/weeks to satisfy condition B
### or PCA the countries for dimension reduction! <ASK PROF: what do the PC's even mean?> youtube.com/watch?v=VjsO5bLe4yg
  # does PCA count as a model? or do we just use it to select the most important variables?
  # analyze countries/regions separately from each other?
### data is specifically the # or % of new cases
  # analyze countries with negative values separately? what do we do?
## time series model? towardsdatascience.com/the-complete-guide-to-time-series-analysis-and-forecasting-70d476bfe775
    ## CHANGE THE QUESTION that we want to answer, in order to avoid needing time series analysis! <?>

# We have (2) dimensions of (419+each_day/188+extra_countries) predictors we can analyze: date, country
# We have (1) dimension of (4) outcomes we can analyze: % new/total cases/deaths
# I'm thinking, if we do PCA, we can have "countries" as PCs and "% new cases" as the data, kind of like how we did PCA for babynames. Either way, can we only decrease one dimension of variables (e.g. which countries/dates contribute most)? Or can we PCA two dimensions of variables at once (e.g. which countries AND dates contribute most)? 
  ## maybe find which days/countries were most severe, and then do further analysis? could this just end up finding the max of each? FIND LOCKDOWN INFO!
```

### ANOVA Model

For the second question of interest, we will be using a two-way ANOVA model to determine how individual policies, particularly facial coverings and stay-at-home restrictions, drafted by governments affect the spread of COVID-19 and thus, the number of new cases. The two-way ANOVA model will be used to examine how two independent variables affect a single dependent variable. In this case, facial coverings and stay-at-home restrictions are independent from each other as one policy applies restrictions when outside while the other applies restrictions when inside, so in theory, we are assuming that they are mutually exclusive. Other assumptions are that there is a linear relationship between the predictor and outcome variables, the variables are independent from each other, no multicollinearity, and the residuals are linear and normally distributed. This model is ideal in understanding not only which policies directly impact the number of new cases, but which levels of those policies are the most significant as well, making it more suitable for our question of interest than regression models that only provide an overall relationship between predictor variables and an outcome variable.

The two-way ANOVA model is defined as $$Y_{ijk}=μ...+α_i+β_j+ϵ_{ijk}$$ where i represents the level of facial covering policy, j represents the level of stay-at-home policy, and μ represents the overall mean of new cases. In addition, $α_i$ represents the effect due to the i-th facial covering policy, $β_j$ represents the effect due to the j-th stay-at-home policy, and $ϵ_{ijk}$ represents the error term in the i-th facial covering policy, j-th stay-at-home policy, and k-th observation. Lastly, $Y_{ijk}$ represents the effect due to interaction between the i-th facial covering policy, j-th stay-at-home policy, and k-th observation.

```{r, echo=FALSE}

#head(policy.data)

COVID.policy.cases.aov = aov(New_cases~facial_coverings+stay_home_requirements,data=policy.data)
summary(COVID.policy.cases.aov)

TukeyHSD(COVID.policy.cases.aov, "facial_coverings")
plot(TukeyHSD(COVID.policy.cases.aov, "facial_coverings"))

TukeyHSD(COVID.policy.cases.aov, "stay_home_requirements")
plot(TukeyHSD(COVID.policy.cases.aov, "stay_home_requirements"))

```

## 5. Model Fitting, & Analysis
<!--Fit the proposed model in (4) and explain your results.-->

### Linear Model

To evaluate whether there is a difference in current mortality rate given the various proposed predictors, we performed an overall F-test as well as individual F-tests. The null hypothesis for this model is $H_0: \beta_{obesity} = \beta_{elderly} = \beta_{HCspent} = \beta_{AQI} = \beta_{density} = \beta_{med.wealth} = \beta_{HDIadj} = 0$, and the alternative hypothesis for this model is $H_a$: ≥1 of the parameters is not 0.

In the process of fitting a model, we found many countries that had missing data or were influential observations that needed to be removed. Malta has a population density of 1634 (which was four times higher than the next most densely populated country), which made further analysis on population density inconclusive. Switzerland has a median wealth of over $220,000 USD, which also lies well beyond other observations - the next-biggest outlier is \$60,000 lower than this observation. In both of these cases, the statistic is well beyond the 1.5\*IQR (or even 2\*IQR) range for outliers, so we were comfortable with removing them from the dataset. 

I used both PCA and pairwise plots to detect collinearity between variables, and avoid violating some linear model assumptions. The variables HCSpentPerCap, HDIadj, and med.wealth were clustered fairly closely together when plotting the PC1 & PC2 values, and therefore accounted for similar variances, so I removed the two less-significant variables and only kept HCSpentPerCap. This is also shown in the pairwise plots, where the three variables have noticeably linear relationships with one another. 

The final fitted model is as follows: $$Y_i = -0.0580 + 0.00094X_{obesity,i} + 0.00195X_{elderly,i} + 0.0000012X_{HCspent,i} + 0.00112X_{AQI,i} + 0.0000110X_{density,i} + \epsilon_{i}$$ where $i$ denotes the $i$'th country observed, and $X_{predictor}$ is the vector of observed values for a specific predictor. 

This model says that, an increase of 1 in any of the predictors (given that the others are fixed) will increase the mortality rate by that predictor's coefficient (e.g. an increased elderly population by one percent, with the others remaining the same, is associated with an increased mortality rate by 0.00195).

The overall F-test for this model has an F-statistic of $8.397$, which corresponds to a p-value of $6.83*10^{-5}$. This is statistically significant at the level $\alpha=0.05$, so we must reject the null hypothesis, which states that all of the parameters are 0; and accept the alternative hypothesis, which states that at least one of the parameters is not 0. 

At the significance level $alpha=0.05$, we must reject the null hypothesis (all coefficients are zero) and accept the alternative hypothesis (at least some coefficients != 0). Further partial F-tests show that higher proportions of elderly people in a population (F = 13.53, p-value = 0.0010), as well as increases in the air quality index (F = 21.80, p-value = 0.000074) and the amount of money spent on healthcare per capita (F = 4.52, p-value = 0.043), are all associated with higher national COVID-19 mortality rates at the significance level $\alpha=0.05$. Elderly populations are at higher risk of dying from COVID-19, as are individuals who have suffered increased lung damage from airborne particulates. The increased mortality rate among countries with higher healthcare spending per capita may be due to the country's healthcare system being less efficient at allocating resources, or the healthcare system already operating near capacity even before the COVID-19 pandemic, and becoming overwhelmed as more patients and staff require medical attention and new equipment. 

```{r, echo=FALSE}
#dont use CNew, DNew yet because the day-to-day variation is too high, and the ways it can change are too numerous. we want to look at how things that happened before the pandemic even started are affecting the outcome
#### We will use mortality rate as the outcome variable because it doesn't just go up all the time, instead staying at a specific rate that can only go up or down very slightly. TODO: add "change in mortality" as a variable?

#preCOVIDtoCNew
#preCOVIDtoCTotal
#preCOVIDtoDNew

#preCOVIDtoDTotal = lm(prcDTotal~ObesityPrc+ElderlyPrc+HCSpentPerCap+AQI+km.density+med.wealth+HDIadj, data=europe.highest.DTotal); summary(preCOVIDtoDTotal); anova(preCOVIDtoDTotal)  #

## MORTALITY RATE MODEL W/EXTRA VARIABLES
preCOVIDtoMort = lm(mortality.now~ObesityPrc+ElderlyPrc+HCSpentPerCap+AQI+km.density+med.wealth+HDIadj, data=europe.recent); # summary(preCOVIDtoMort); anova(preCOVIDtoMort)  #


## TRY REMOVING MALTA (outlier): 
withoutmaltaEtc = europe.recent[-c(28, 40),]  # removed malta, maybe also netherlands/belgium for density, switzerland for income, as outliers?
mortnoMaltaEtc = lm(mortality.now~ObesityPrc+ElderlyPrc+HCSpentPerCap+AQI+km.density+med.wealth+HDIadj, data=withoutmaltaEtc); # summary(mortnoMaltaEtc); anova(mortnoMaltaEtc)

#plot(europe.recent$ObesityPrc, europe.recent$mortality.now)
#plot(europe.recent$ElderlyPrc, europe.recent$mortality.now)
#plot(europe.recent$HCSpentPerCap, europe.recent$mortality.now)
#plot(europe.recent$AQI, europe.recent$mortality.now)
par(mfrow=c(2, 2))
## DEMONSTRATING MALTA AS OUTLIER
plot(europe.recent$km.density, europe.recent$mortality.now, main="Pop. Dens., incl. Malta", ylab="Mortality Rate", xlab="Pop. Density")
plot(withoutmaltaEtc$km.density, withoutmaltaEtc$mortality.now, main="Pop. Dens., w/o Malta", ylab="Mortality Rate", xlab="Pop. Density")
## DEMONSTRATING SWITZERLAND AS OUTLIER
plot(europe.recent$med.wealth, europe.recent$mortality.now, main="Wealth, incl. Switzerland", ylab="Mortality Rate", xlab="Wealth, per capita")
plot(withoutmaltaEtc$med.wealth, withoutmaltaEtc$mortality.now, main="Wealth, w/o Switzerland", ylab="Mortality Rate", xlab="Wealth, per capita")
#plot(europe.recent$HDIadj, europe.recent$mortality.now)
par(mfrow=c(1, 1))

## NEW DATASET
woMaltafiltered = withoutmaltaEtc %>% filter(!is.na(AQI))  # AQI has the most missing values, and this gets rid of some other outlier countries
## CHECKING FOR COLLINEARITY
    pairs(cbind(woMaltafiltered$mortality.now,woMaltafiltered$ObesityPrc,woMaltafiltered$ElderlyPrc,woMaltafiltered$HCSpentPerCap,woMaltafiltered$AQI,woMaltafiltered$km.density,woMaltafiltered$med.wealth,woMaltafiltered$HDIadj), labels=c("Mortality", "Obesity", "Elderly", "HC. Spent", "AQI", "Pop. Dens.", "Med. Wealth", "HDI adj."), main="Pairwise Plots, w/o NA values/countries", gap=0.15, pch=3)
## PRINCIPAL COMPONENT ANALYSIS FOR VARIANCE/EIGENVECTORS
eu.pca = select(woMaltafiltered, c("ObesityPrc", "ElderlyPrc", "HCSpentPerCap", "AQI", "km.density", "med.wealth", "HDIadj")) %>% prcomp(center=TRUE, scale=TRUE)
## SCREE PLOT
plot(eu.pca, type="lines", main="Scree Plot for the PCA")
    "First Two PC's Coefficients: "
    eu.pca[["rotation"]][,1];eu.pca[["rotation"]][,2]
pca.frame = data.frame(cbind(country=woMaltafiltered$Country, unclass(eu.pca$x)))
    #plot(pca.frame$PC1, pca.frame$PC2); abline(v=0); abline(h=0)
## VISUALIZING PC CONTENTS
    plot(eu.pca[["rotation"]][,1], eu.pca[["rotation"]][,2], xlab="PC1",ylab="PC2", main="First Two PCs' Transformations", pch=9); abline(v=0); abline(h=0) # new axes
    text(eu.pca[["rotation"]][,2]~eu.pca[["rotation"]][,1], labels=row.names(eu.pca[["rotation"]]), adj=c(-0.2, 1),cex=0.55, srt=15)
# woMfilt.long = pivot_longer(pca.frame, cols=!country, names_to="PC", values_to="Coefficient")
# woMfilt.long %>% filter(PC %in% c('PC1','PC2','PC3','PC4','PC5','PC6','PC7')) %>% ggplot(aes(x=country, y=Coefficient, colour = PC)) + geom_line()

## FINAL MODEL, AFTER REMOVING EXTRA VARS
final.mortnoMaltaEtc = lm(mortality.now~ObesityPrc+ElderlyPrc+HCSpentPerCap+AQI+km.density, data=withoutmaltaEtc); #summary(final.mortnoMaltaEtc); anova(final.mortnoMaltaEtc)


# moving average model (dont use this yet! define a good question that we can make a good model for! few confounders, etc.)
# https://financetrain.com/estimating-moving-average-ma-model-in-r/
#US_MA = arima(x=UScovid$prcCNew)  #another var?
#US_MA; summary(US_MA)
#ts.plot(UScovid$prcCNew)
#resids = US_MA$residuals
#ts.plot(resids)
```

### ANOVA Model

To evaluate whether there is a difference in new cases among different levels of facial covering mandates and stay-at-home restrictions, an F-test was performed. The null hypothesis for this model is that $H_0: \alpha_1=\alpha_2=\alpha_3=0$, meaning that there is no difference in mean new cases among different levels of policies, while the alternative hypothesis is Ha: not all \alpha s are zero, meaning there is a difference in mean new cases among different levels of policies. After calculating the summary of our fit ANOVA mode, the test statistic for facial coverings was 67.94 with a P-value of 2e-16, while the test statistic for stay-at-home requirements was 4.78 with a P-value of 0.00934. Since both of these P-values are much lower than the significance level of 0.05, we reject the null hypothesis and conclude that there exists a difference in mean new cases among different levels of policies for facial coverings and stay-at-home requirements. 

Tukey’s range test was used to determine which level of each policy was associated with the highest difference in new cases, with a plot of each policy’s 95% confidence interval generated as well. The assumptions of the test are that the groups associated with each mean are normally distributed, homogeneity of variance is present, and the tested observations are independent for the groups. Based on these results, the difference in the number of new cases with facial coverings recommended is higher than that of facial coverings required in some specified shared or public spaces by roughly 514.6282 to 2564.889 new cases with an average of 1539.7588 cases. Additionally, this difference is higher for recommended facial coverings than coverings required outside the home at all times by -2226.5205 to 1337.206 cases with an average of 444.6571 cases. Facial coverings required outside at all times also results in a higher difference than coverings required in some public spaces by 4579.671 to 6721.2047 cases with an average of 5650.4381, in addition to coverings required in all public spaces by 3399.339 to 7012.2229 cases with an average of 5205.7810 cases. 

Since these ranges do not overlap with zero and each have P-values much smaller than 0.05, then we can conclude with 95% confidence that requiring facial coverings outside at all times results in the least number of new cases compared to other policies. Additionally, based on these results, the difference in the number of new cases did not significantly vary with different levels of stay-at-home restrictions, so the null hypothesis cannot be rejected for this policy and stay-at-home restrictions did not impact the number of new cases. 


## 6. Model Diagnostics
<!--Conduct model diagnostics and/or sensitivity analysis. -->

### Linear Model

Based on the fitted-to-residuals plot, the residual terms $\epsilon_i$ seem to have mostly constant variance as the x-variables change (with a few exceptions), but there is some slight non-linearity in the residuals as well as some outliers (specifically Hungary, Serbia, and Slovakia). The errors seem to be generally normally distributed according to the QQ-plot. The Shapiro-Wilk test for normality has a p-value of 0.7393, which is not statistically significant at the level $\alpha = 0.05$, so it seems reasonable to accept the null hypothesis: the errors are normally distributed. 

The aforementioned outliers also make the scale-location plot somewhat nonlinear, and there are some outliers in terms of Cook's distance (including an additional outlier: Bosnia & Herzegovina) and the data's leverage. 

The variance inflation factor has a value of 5.613 (≥5) for HCSpentPerCap when all of the variables from the full model are analyzed, so there is some multicollinearity between the variables. After removing some similar variables (as determined by clusters in the PCA), the VIF has decreased enough to reasonably assume a lack of multicollinearity. 

```{r, echo=FALSE}
par(mfrow=c(2, 2))
plot(final.mortnoMaltaEtc, which=1)
plot(final.mortnoMaltaEtc, which=2)
plot(final.mortnoMaltaEtc, which=3)
plot(final.mortnoMaltaEtc, which=5)
par(mfrow=c(1, 1))
shapiro.test(resid(final.mortnoMaltaEtc))
"VIF for original & final models"
vif(mortnoMaltaEtc)
vif(final.mortnoMaltaEtc)
```

### ANOVA Model

Based on the residuals versus fitted plot, the residuals seem to be equally spread due to the red horizontal line around the y = 0 line, indicating that there is a linear relationship between the two values with equal variance of errors with minimal outliers. However, according to the normal Q-Q plot, the generated plot does not have a completely straight line across the residuals as there are heavy tails on either end and the slope is relatively small, indicating that the residuals are not normally distributed and further analysis is required. A scale-location plot was also generated, which indicates that the residuals are not equally spread and that there is indeed unequal variance among these points, so further analysis is also required. Lastly, the residuals versus leverage plot has a horizontal line around y = 0, so it seems that there are no outlier points contributing to the linear relationship. As a result, some of the assumptions for the classic linear model such as the error term having constant variance may not have been met.

##### Residuals Vs Fitted

```{r, echo=FALSE}
plot(COVID.policy.cases.aov,which=1)
```

##### Normal Q-Q

```{r, echo=FALSE}
plot(COVID.policy.cases.aov, which=2)
```

##### Scale-Location

```{r, echo=FALSE}
plot(COVID.policy.cases.aov, which=3)
```

##### Residuals vs Leverage

```{r, echo=FALSE}
plot(COVID.policy.cases.aov,which=5)
```

## 7. Conclusion
<!--Conclude your analysis with a discussion of your findings and caveats of your approach. -->

### Linear Model

From our analysis, we saw that the higher a country's air pollution levels, healthcare spending per capita, and elderly proportion of the population, the higher the mortality rate they were likely to report. One interesting possibility based on this analysis is that how affluent a country is, measured through median wealth and adjusted HDI, still cannot fully account for the variance in a country’s COVID-19 mortality rate: these predictors do not account for as much variance as healthcare expenditures per capita does, which may imply that a good healthcare system (perhaps exemplified through efficiently spending money/acquiring resources and not being near full capacity, where scarcity will drive up costs) may be a stronger predictor of mortality rate than the wealth and development of the country. It may also be possible that wealth and development prevent COVID-19 transmission (due to the possibility of working from home and buying PPE) but not deaths (which probably occur once a patient is already in the healthcare system), so further analysis is necessary, as well as some method of normalizing wealth with cost of living. 

A major caveat of this approach is that we focused solely on larger European countries, specifically comparing Eastern to Western European countries; such a distinct divide does not officially exist within Europe, especially as many Eastern European countries have begun to assimilate into greater Europe and the EU since the end of the Warsaw Pact. As for air pollution levels and healthcare practices, a caveat of this approach would be that we did not compare Europe as a whole or other first-world countries, such as Canada, to a higher-polluted area with different healthcare practices, such as China or India (which are both exceptionally populated and polluted due to their status as a developing country). Furthermore, the presence of air quality monitoring systems were mostly in cities; though this may account for the fact that more people generally live in cities, it is not fully adjusted for the population density that any given individual could live in. However, we were still able to establish that exposure to higher air pollution levels on a national scale does exacerbate a population’s vulnerability to the effects of coronavirus. 

Another caveat of our approach is that Italy's elderly population has, for a long time, been a large percentage of their country's entire population as do other smaller Eastern European countries. By specifically concentrating on average population age within Europe, our approach may be skewed in that we saw a higher average population age than we would have seen in Mexico or Canada. Furthermore, we noticed similar safety practices and guidelines enacted across the continent (likely due to the European Union) versus what we may have seen if we were to compare Europe to other countries (e.g. the United States) in later analysis on policy changes throughout the last year. 
 
### ANOVA Model (Policy changes through March - October)

Based on the analysis above, we can conclude that requiring facial coverings outside at all times has the greatest impact on reducing the number of new cases compared to other tiers, while stay-at-home restrictions did not seem to impact the number of new cases as much, which is contradictory to common belief. However, a major caveat includes the fact that there exists multicollinearity between the variables for the two individual policies, so independent analysis of both variables is required to properly analyze the model and interpret its results. Another alternative would be to choose another policy entirely in order to avoid interaction between the predictor variables. For example, future research can be conducted using facial coverings and testing policies instead, as they are more independent from each other than the currently chosen policies. A second caveat is that the two-way ANOVA model for these policies was only fit to the number of new cases in France. Future research can be conducted by aggregating multiple countries that are similar to one another in terms of how they implement these policies, and performing a two-way ANOVA model with the aggregated countries and a single policy as the predictor variables. A third caveat to our approach is that the timeframe for our analysis is limited to a six month period. Due to the major holiday season, many countries experienced spikes in cases, and therefore made changes to their existing policies by either enforcing them more strictly (i.e. enforcing lockdowns) or relaxing them (i.e. ending curfews). We limited our data analysis to the months before the holiday season to avoid any variance in the fluctuating number of cases and extreme policy changes that took place after October.

## Appendix: Code, Citations

### Code

```{r, ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}

```

### Citations (data, then code)

##### Data: 

Andorra Demographics Profile, www.indexmundi.com/andorra/demographics_profile.html. 

Anonymous. “Country Health Profiles.” Public Health - European Commission, 7 Feb. 2020, ec.europa.eu/health/state/country_profiles_en. 

Authors Jonathan Lehne Assistant Professor | SITE, et al. “Pollution and the COVID-19 Pandemic: Air Quality in Eastern Europe • FREE NETWORK.” FREE NETWORK, 1 Feb. 2021, freepolicybriefs.org/2021/02/01/air-quality-eastern-europe/#:~:text=Figure%201%2C%20which%20plots%20these,to%20their%20Western%20European%20counterparts. 

Culbertson, Alix. “COVID-19: What Are the Christmas Coronavirus Rules in Other European Countries?” Sky News, Sky, 19 Dec. 2020, news.sky.com/story/covid-19-what-are-europes-christmas-coronavirus-rules-12162445. 

“Coronavirus Disease 2019.” Wikipedia, Wikimedia Foundation, 5 Mar. 2021, en.wikipedia.org/wiki/Coronavirus_disease_2019. 

“Countries Ordered by Population in 2019.” PopulationPyramid.net, www.populationpyramid.net/population-size-per-country/2019/. 

“Countries With the Oldest Populations in the World.” Population Reference Bureau, 16 Mar. 2019, www.prb.org/countries-with-the-oldest-populations/. 

“EU Countries - The Member States of the European Union.” SchengenVisaInfo.com, 24 Nov. 2020, www.schengenvisainfo.com/eu-countries/.

“FACTOR VARIABLES | R LEARNING MODULES.” IDRE Stats, University of California, Los Angeles, stats.idre.ucla.edu/r/modules/factor-variables/. 

“CODING FOR CATEGORICAL VARIABLES IN REGRESSION MODELS | R LEARNING MODULES.” IDRE Stats, University of California, Los Angeles, https://stats.idre.ucla.edu/r/modules/coding-for-categorical-variables-in-regression-models/.

“Fine Particulate Matter and COVID-19 Mortality in the United States.” Home, projects.iq.harvard.edu/covid-pm. 

“Is Air Quality in China a Social Problem?” ChinaPower Project, 26 Feb. 2021, chinapower.csis.org/air-quality/.
“List of Countries and Dependencies by Population Density.” Wikipedia, Wikimedia Foundation, 27 Feb. 2021, en.wikipedia.org/wiki/List_of_countries_and_dependencies_by_population_density#cite_ref-9. 
“List of Countries by Age Structure.” Wikipedia, Wikimedia Foundation, 15 Feb. 2021, en.wikipedia.org/wiki/List_of_countries_by_age_structure. 

“List of Countries by Life Expectancy.” Wikipedia, Wikimedia Foundation, 24 Feb. 2020, en.wikipedia.org/w/index.php?title=List_of_countries_by_life_expectancy&oldid=942400940#List_by_the_World_Health_Organization_(2015). 

“List of Countries by Inequality-Adjusted HDI.” Wikipedia, Wikimedia Foundation, 1 Mar. 2021, en.wikipedia.org/wiki/List_of_countries_by_inequality-adjusted_HDI. 

“List of Countries by Total Health Expenditure per Capita.” Wikipedia, Wikimedia Foundation, 6 Feb. 2021, en.wikipedia.org/wiki/List_of_countries_by_total_health_expenditure_per_capita#World_Health_Organization. 

“List of Countries by Wealth per Adult.” Wikipedia, Wikimedia Foundation, 3 Feb. 2021, en.wikipedia.org/wiki/List_of_countries_by_wealth_per_adult#By_country. 

“Member State of the European Union.” Wikipedia, Wikimedia Foundation, 5 Mar. 2021, en.wikipedia.org/wiki/Member_state_of_the_European_Union#List. 

Ouyang, Xiao, et al. “Environmental Regulation, Economic Growth and Air Pollution: Panel Threshold Analysis for OECD Countries.” Science of The Total Environment, Elsevier, 5 Dec. 2018, www.sciencedirect.com/science/article/pii/S0048969718348964. 

“Population Ages 65 and above (% of Total Population).” Data, data.worldbank.org/indicator/SP.POP.65UP.TO.ZS?name_desc=false&view=chart. 

Ritchie, Research and data: Hannah. “COVID-19: Cancellation of Public Events and Gatherings.” Our World in Data, ourworldindata.org/covid-cancel-public-events. 

Ritchie, Research and data: Hannah. “COVID-19: Face Coverings.” Our World in Data, ourworldindata.org/covid-face-coverings. 

Ritchie, Research and data: Hannah. “COVID-19: Stay-at-Home Restrictions.” Our World in Data, ourworldindata.org/covid-stay-home-restrictions. 

Spiteri, Gianfranco, et al. “First Cases of Coronavirus Disease 2019 (COVID-19) in the WHO European Region, 24 January to 21 February 2020.” Eurosurveillance, European Centre for Disease Prevention and Control, 5 Mar. 2020, www.eurosurveillance.org/content/10.2807/1560-7917.ES.2020.25.9.2000178. 

Tandon, Ajay, et al. “MEASURING OVERALL HEALTH SYSTEM PERFORMANCE FOR 191 COUNTRIES.” World Health Organization, World Health Organization, www.who.int/healthinfo/paper30.pdf.

“Two-Way ANOVA in SPSS Statistics.” Two-Way ANOVA in SPSS Statistics - Step-by-Step Procedure Including Testing of Assumptions | Laerd Statistics, statistics.laerd.com/spss-tutorials/two-way-anova-using-spss-statistics.php. 

***

##### Code: 

grouping one variable by highest of another variable - https://stackoverflow.com/questions/24558328/how-to-select-the-row-with-the-maximum-value-in-each-group

removing rows that are duplicates by one variable - https://www.datanovia.com/en/lessons/identify-and-remove-duplicate-data-in-r/

concatenate/print strings - https://www.math.ucla.edu/~anderson/rw1001/library/base/html/cat.html

add text to plots - https://rpubs.com/RatherBit/188960

hiding unnecessary method outputs - https://stackoverflow.com/questions/52433164/how-do-i-remove-the-code-on-column-specification-in-r-markdown-while-preserving

StatQuest on youtube for PCA - https://www.youtube.com/watch?v=FgakZw6K1QQ

Drop data frame columns by name - https://stackoverflow.com/questions/4605206/drop-data-frame-columns-by-name
